DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2021-11-04 manuscript/manuscript.Rmd
some tweaks to the comparison with timetree and a typo fix

Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database. This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`. Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining [@Kumar2017].
In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015]. This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and is part of datelife's database now. Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage. Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
<!--
Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services. One option to increase our database is the Dryad data repository.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
<!-- Can we mine journal databases? -->
<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OpenTree's database. -->
Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository [@mctavish2015phylesystem].

@@

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses. For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations. Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence). Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms. Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not-->
However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis. Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram. <!-- Example -->

>>>>

Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database. This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`. Unfortunately, TimeTree's database is not open for scientific reuse nor automated data mining [@Kumar2017].
In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015]. This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and it is part of datelife's database now. Hence, the amount of lineages represented in datelife's database (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa), is at least as substantial as TimeTree's (as of the last released tree, with `r max(sapply(opentree_chronograms$trees[grepl("Hedges", names(opentree_chronograms$trees))], ape::Ntip))` taxa). Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
<!--
Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services. One option to increase our database is the Dryad data repository.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
<!-- Can we mine journal databases? -->
<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OpenTree's database. -->
Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository [@mctavish2015phylesystem].

@@

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses. For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations. Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence). Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms. Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not-->
However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis. Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram. <!-- Example -->
